---
title: "04_Advanced_Visualizations"
author: "Jack Chiang, Louis Lax-Roseman"
date: "2025-04-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

1. Introduction

In this document we generate advanced visualizations to explore transcription factor (TF) family distributions and expression across species. The analyses include:

    Clustered stacked bar plots for TF family distributions.

    Normalized heatmaps of TF family abundance.

    An improved PCA biplot with loadings.

    A correlation network of TF families.

    Hierarchical clustering heatmaps.
  

2. Clustered Stacked Bar Plot for TF Family Distribution

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(colorspace)


```

Load TF family summary data (adjust file path as needed):

```{r}
tf_summary <- read.delim("data/tf_family_summary.tsv", sep = "\t", header = TRUE)

```

Pivot data into long format for plotting:

```{r}
stacked_data <- tf_summary %>%
  select(TF_Family, Cre, Mpo, Ppa, Smo, Pab, Ath, Zma) %>%
  pivot_longer(cols = c(Cre, Mpo, Ppa, Smo, Pab, Ath, Zma),
               names_to = "Species",
               values_to = "Gene_Count")

```

Generate a custom color palette:

```{r}
n_colors <- 15
my_palette <- rainbow_hcl(n_colors, c = 60, l = 70)
my_palette <- sample(my_palette, n_colors)

```
Create the stacked bar plot:

```{r}
ggplot(stacked_data, aes(x = Species, y = Percentage, fill = TF_Family)) +
  geom_bar(stat = "identity") +
  facet_grid(. ~ Species, scales = "free_x") +
  theme_minimal() +
  labs(title = "Clustered Stacked Bar Plot: TF Family Distribution",
       x = "Species", y = "Percentage (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = my_palette)

```

3. Normalized Heatmap of TF Family Abundance

```{r}
library(pheatmap)
library(viridis)
library(tibble)


```
Assume tf_summary has TF_Family and species columns (Cre, Mpo, Ppa, Smo, Pab, Ath, Zma):

```{r}
species_cols <- c("Cre", "Mpo", "Ppa", "Smo", "Pab", "Ath", "Zma")
species_totals <- colSums(tf_summary[, species_cols])

norm_data <- tf_summary %>%
  select(TF_Family, all_of(species_cols)) %>%
  column_to_rownames("TF_Family")

for(col in species_cols) {
  norm_data[, col] <- norm_data[, col] / species_totals[col] * 100
}

```
Create the normalized heatmap:
```{r}
pheatmap(as.matrix(norm_data),
         display_numbers = TRUE,
         number_format = "%.1f%%",
         color = viridis(100),
         main = "Normalized TF Family Abundance Across Species",
         fontsize = 12,
         fontsize_number = 10)

```

4. Improved PCA Biplot for TF Family Expression
```{r}
library(ggrepel)


```

Use tf_summary for PCA (rows = TF families, columns = species counts):
```{r}
pca_data <- tf_summary %>%
  select(TF_Family, all_of(species_cols)) %>%
  column_to_rownames("TF_Family") %>%
  t()

pca_result <- prcomp(pca_data, scale. = TRUE)
pca_df <- as.data.frame(pca_result$x)
pca_df$Species <- rownames(pca_df)


```
Extract loadings and calculate contributions:
```{r}
loadings <- pca_result$rotation
loadings_df <- as.data.frame(loadings)
loadings_df$TF_Family <- rownames(loadings_df)
loadings_df$PC1_contrib <- loadings_df$PC1^2
loadings_df$PC2_contrib <- loadings_df$PC2^2
loadings_df$total_contrib <- loadings_df$PC1_contrib + loadings_df$PC2_contrib

# Filter loadings for plotting
loadings_plot <- loadings_df %>% filter(PC1_contrib > 0.03 | PC2_contrib > 0.03)
loadings_plot$is_strong <- loadings_plot$total_contrib > median(loadings_plot$total_contrib)
scaling_factor <- 6


```

Base PCA plot:
```{r}
p1 <- ggplot() +
  geom_point(data = pca_df, aes(x = PC1, y = PC2, color = Species), size = 4) +
  geom_text_repel(data = pca_df, aes(x = PC1, y = PC2, label = Species), size = 3) +
  theme_minimal() +
  labs(x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)"),
       title = "PCA Biplot of TF Family Expression")

```

Add arrows for loadings:
```{r}
p_biplot <- p1 +
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = PC1 * scaling_factor, yend = PC2 * scaling_factor, size = total_contrib),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "darkblue") +
  scale_size_continuous(range = c(0.5, 2), guide = "none") +
  geom_text_repel(data = loadings_plot,
                  aes(x = PC1 * scaling_factor, y = PC2 * scaling_factor, label = TF_Family),
                  size = 3,
                  box.padding = 0.5,
                  color = "black")
p_biplot


```



5. Correlation Network of TF Families

```{r}
library(igraph)
library(ggraph)

```
Calculate correlation between TF families across species using the transposed normalized data:
```{r}
tf_cor <- cor(t(as.matrix(norm_data)), method = "spearman")
rownames(tf_cor) <- rownames(norm_data)
colnames(tf_cor) <- rownames(norm_data)

# Remove self-correlations and set a threshold to filter weak correlations
tf_cor[abs(tf_cor) < 0.6] <- 0
diag(tf_cor) <- 0

# Build a graph object from the correlation matrix
graph <- graph_from_adjacency_matrix(tf_cor, mode = "undirected", weighted = TRUE)
E(graph)$width <- abs(E(graph)$weight) * 5
E(graph)$color <- ifelse(E(graph)$weight > 0, "blue", "red")

set.seed(123)  # For reproducibility
plot(graph, vertex.size = 20, vertex.label.color = "black",
     vertex.color = "lightblue", vertex.frame.color = "gray",
     vertex.label.cex = 0.8, edge.curved = 0.2, layout = layout_with_fr(graph))


```

6. Hierarchical Clustering Heatmap

Save the clustered heatmap as a PNG file:

```{r}
png("TF_combined_clustered_heatmap.png", width = 1200, height = 1000, res = 150)
pheatmap(as.matrix(norm_data),
         display_numbers = TRUE,
         number_format = "%.1f%%",
         color = viridis(100),
         main = "Hierarchical Clustering of TF Families and Species",
         fontsize = 12,
         fontsize_number = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2")
dev.off()


```

# References
- Kolde, R. (2019). pheatmap: Pretty Heatmaps. R package version 1.0.12.
- Jolliffe, I. T. (2002). Principal Component Analysis. Springer.
- Csardi, G. & Nepusz, T. (2006). The igraph software package for complex network research.

